---
title: "Introduction to Classification"
output: html_notebook
---

```{r,warning=FALSE,message=FALSE}
library(tidyverse)
library(broom)
library(caret)
```


# Introduction 

Recall the iris data set:
```{r}
head(iris)
```

Attempting to predict the species of a flower based on the length and width measurements is a **classification** problem. In this case, the reponse variable assumes one of three possible class values: setosa, virginica, or versicolor. 

There is a special type of classification problem, **binary classification** where the categorical response variable takes values from just two distinct possible outcomes. There is a standard regression method for binary classification called logistic regression. Since regression is already a familiar concept, we will begin by discussing logistic regression, how it is implemented in R, and how to apply it. 

# Logistic Regression


# KNN for Classification

This [online demo](https://codepen.io/gangtao/pen/PPoqMW) is very helful. 
